---
title: "Examples"
author: "Thomas Laetsch"
date: "October 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Two Extra Examples

Although we have already included three examples in the README file, here we provide two more special case examples and how to deal with them. 

```{r source }
source( 'MonthlyHomCnts.R' )
```

### Example 1: xlsx file

We start with the file `Minneapolis_homicide_counts.xlsx` in the `input_files` directory, which is in xlsx format. Such a file cannot be immediately read into the function `format_monthly_hom_cnts` and first needs to be converted to csv. With the package `xlsx` installed, converting to a csv is quite simple.

```{r convert_xlsx_to_csv}
library(xlsx)
xlsx_fp <- 'input_files/Minneapolis_homicide_counts.xlsx'
xlsx_df <- read.xlsx( xlsx_fp, sheetIndex = 1 )
head( xlsx_df )
```
Note that we set `sheetIndex = 1`, which is because the data was on the first sheet of the Excel notebook (if this is not the case, choose the appropriate sheet index). Now, save as csv:
```{r save_xlsx_df_as_csv }
csv_fp <- 'input_files/Minneapolis_hom_cnts.csv'
write.csv( x = xlsx_df, file = csv_fp, row.names = F )
```
where we set `row.names = F` since we don't need nor want the output csv to contain a column of row indices. From here, we're ready to move on to formatting as before. 
```{r xlsx_to_csv_format }
df <- format_monthly_hom_cnts( csv_fp )
save_monthly_df( df, output_dir = 'output_files/' )
```

### Example 2: Deliminated other than csv

The next example will use the file `HPD_Monthly_Homicide_Count.csv` in the `input_files` directory. Although labelled as `csv`, the errors thrown trying to run it through `format_monthly_hom_cnts` inspired us to look at the original file.
```{r look_at_HPD}
tsv_fp <- 'input_files/HPD_Monthly_Homicide_Count.csv'
HPD_df <- read.csv( tsv_fp, stringsAsFactors = F )
head(HPD_df)
```
From this, we see that it was read in with only one long, period-concatenated column label, and that the entry has the information partitioned by `\t`, suggesting these entries were tab delimited. The function `format_monthly_hom_cnts` can handle this, but must be told explicitly that the separation is tab, rather than comma; we do this by adding the argument `sep='\t'`. Before jumping into this, we also notice that neither the city name nor the state are included, so we need to send these arguments as well. 
```{r tab_sep_HPD}
df <- format_monthly_hom_cnts( tsv_fp, place_name = 'Houston', state_abr = 'TX', sep = '\t' )
save_monthly_df( df, output_dir = 'output_files/' )
```